NumLin: Linear Types for Linear Algebra

### 2019-04-13 write-up/paper/formal_system.tex
Make stuff consistent
Syntax & spacing, with a few small edits from Reviewer C's review.

The \highl{Many} introduction and elimination rules are very important.
Producing !-type values may only be done if the expression inside is a
syntactic value which is not a location. This allows all safely duplicable
resources, including functions which capture non-linear resources from their
environments, but prevents producing aliases of (pointers to) arrays and
matrices. This is exactly the same as value-restriction from the world of
parametric polymorphism.

@@

Lastly, types of almost all \lang\ primitives, as embedded in OCaml's type
system, are shown in Appendix \ref{subsec:primitives}, with some similar ones
(like those for binary arithmetic operators) omitted for brevity. The
main difference between the OCaml type of a primitive like \highl{gemm} and its
\lang\ counterpart is the inclusion of explicit `$\forall$'s.  So,
\highl{float bang -> ('a mat * bool bang) -> ('b mat * bool bang) -> float
bang -> z mat -> ('a mat * 'b mat) * z mat}
will correspond to \\
$!\ottkw{elt} \multimap \forall\, x.\ x \ \ottkw{mat} \ \otimes \ !\ottkw{bool}
\multimap \forall\, y.\  y \ \ottkw{mat} \ \otimes \ !\ottkw{bool} \multimap
\ !\ottkw{elt} \multimap z\ \ottkw{mat} \multimap ( x \ \ottkw{mat} \ \otimes y
\ \ottkw{mat} ) \ \otimes z\ \ottkw{mat}$

@@

We draw the reader's attention to the definitions relating to fractional
permissions. Specifically, unlike a lambda, the body of a $\ottkw{fun}\, f\!c
\rightarrow \_$ must be a syntactic value. The context $\ottkw{fun}\, f\!c
\rightarrow [-]$ means expressions can be reduced inside a fractional
permission generalisation. This is to emphasize that fractions are merely
\emph{compile-time constructs} and do not affect runtime behaviour. Correct
usage of fractions is enforced by the type system, so programs do not get
stuck. Fractional permissions are specialised using substitution over both the
heap and an expression (\textsc{Op\_Frac\_Perm}).

@@

\vspace{-\baselineskip}
\[
    \den{H}{}{\sigma} = \bigstar_{(l,f,m) \in \sigma} [ l \mapsto_f m ]
\]

@@

We then define a step-indexed logical relation in the style of Morrisett et.
al~\cite{morrisett}. $(\varsigma, v) \in \V{k}{t}$ means it takes a heap with
exactly $\varsigma$ resources to produce a value $v$ of type $t$ in at most $k$
steps. So, something like a $\ottkw{unit}$ or a $!t$ need no resources, whereas
a $f\, \ottkw{mat}$ needs exactly $f$ ownership of a some matrix and a pair
needs a $\star$ combination of the heaps required for each component.
\begin{align*}
  \V{k}{ \Unit } &= \{ (\empH, \ast) \} \\
  \V{k}{ f \, \Mat } &= \{ (\{ l \mapsto_{2^{-f}} \_ \} , l) \} \\
  \V{k}{ \Bang t } &= \{ (\empH, \Many\, v) \mid (\empH, v) \in \V{k}{t} \} \\
  \V{k}{ t_1 \otimes t_2 } &= \{ (\varsigma_1 \star \varsigma_2, ( v_1, v_2 )) \mid (\varsigma_1, v_1) \in \V{k}{t_1} \wedge (\varsigma_2, v_2) \in \V{k}{t_2} \}
\end{align*}

The definition of $\V{k}{\forall f\!c.\ t}$ says a value and heap
must be the same regardless of what fraction is substituted into both; the
$k-1$ is to take into account fraction specialisation takes ones step
(\textsc{Op\_Spc}).

\vspace{-\baselineskip}
\[
    \V{k}{ \forall f\!c.\  t } = \{ (\varsigma, \ottkw{fun}\, f\!c \rightarrow \, v) \mid \forall f.\ (\varsigma [ f\!c / f ], v [ f\!c / f ]) \in \V{k-1}{ t [ f\!c / f ] } \}
\]

@@

In this light, $\V{k}{t' \multimap t}$ simply says
that $v$ is a function and that evaluating the application of it to any
argument (of the correct type, requiring its own set of resources, bounded by
$k$ steps) satisfies all the aforementioned properties.
\begin{align*}
    \V{k}{ t' \multimap t } &= \{ (\varsigma_v, v ) \mid ( v \equiv \ottkw{fun}\, x : t' \rightarrow e \vee v \equiv \ottkw{fix} (g, x : t' , e : t) ) \, \wedge\\
                            & \qquad \qquad \forall j \leq k, (\varsigma_{v'}, v') \in \V{j}{ t' }.\ \varsigma_v \star \varsigma_v' \textrm{ defined } \Rightarrow (\varsigma_v \star \varsigma_v', v\, v') \in \C{j}{t} \}
\end{align*}

The interpretation of typing environments $\Delta$ and $\Gamma$ are with
respect to an arbitrary substitution of fractional permissions $\theta$. Note
that only the interpretation of $\Gamma$ involves a (potentially) non-empty heap.
\begin{align*}
    \den{I}{k}{ \Delta, x : t } \theta &= \{ \delta[x \mapsto v_x ] \mid \delta \in \den{I}{k}{\Delta}\theta \wedge (\empH, v_x) \in \V{k}{\theta(t)} \} \\
    \den{L}{k}{ \Gamma, x : t } \theta &= \{ (\varsigma \star \varsigma_x, \gamma[x \mapsto v_x ]) \mid (\varsigma, \gamma) \in \den{L}{k}{\Gamma}\theta \wedge (\varsigma_x, v_x) \in \V{k}{\theta(t)} \}
\end{align*}

>>>>

The \highl{Many} introduction and elimination rules are very important.
Producing !-type values may only be done if the expression inside is a
syntactic value which is not a location. This allows all safely duplicable
resources, including functions which capture non-linear resources from their
environments, but prevents producing aliases of (pointers to) arrays and
matrices. This is exactly the same as value-restriction from the world of
parametric polymorphism; without it, the expression $\Let\ \Many\ x = \Many\
(\ottkw{array}\ 5)\ \In\ \Let\ () = \ottkw{free}\ x\ \In\ x[0]$ would
type-check but error at runtime.

@@

Lastly, types of almost all \lang\ primitives, as embedded in OCaml's type
system, are shown in Appendix~\ref{subsec:primitives}, with some similar ones
(like those for binary arithmetic operators) omitted for brevity. The main
difference between the OCaml type of a primitive like \highl{gemm} and its
\lang\ counterpart (Figure~\ref{fig:primitive_types}) is the inclusion of
explicit universal-quantification of fractional permission variables in the
latter.

@@

We draw the reader's attention to the definitions relating to fractional
permissions. Specifically, unlike a lambda, the body of a $\ottkw{fun}\ f\!c
\rightarrow v$ must be a syntactic value. The context $\ottkw{fun}\ f\!c
\rightarrow [-]$ means expressions can be reduced inside a fractional
permission generalisation. This is to emphasize that fractions are merely
\emph{compile-time constructs} and do not affect runtime behaviour. Correct
usage of fractions is enforced by the type system, so programs do not get
stuck. Fractional permissions are specialised using substitution over both the
heap and an expression (\textsc{Op\_Frac\_Perm}).

@@

\vspace{-\baselineskip}
\[
    \den{H}{}{\sigma} = \bigstar_{(l,f,m) \in \sigma} [l \mapsto_f m]
\]

@@

We then define a step-indexed logical relation in the style of Morrisett et.
al~\cite{ahmed20073}. $(\varsigma, v) \in \V{k}{t}$ means it takes a heap with
exactly $\varsigma$ resources to produce a value $v$ of type $t$ in at most $k$
steps. So, something like a $\ottkw{unit}$ or a $!t$ need no resources, whereas
a $f\, \ottkw{mat}$ needs exactly $f$ ownership of a some matrix and a pair
needs a $\star$ combination of the heaps required for each component.
\begin{align*}
  \V{k}{ \Unit } &= \{ (\empH, \ast) \} \\
  \V{k}{ f \, \Mat } &= \{ (\{ l \mapsto_{2^{-f}} \_ \} , l) \} \\
  \V{k}{ \Bang t } &= \{ (\empH, \Many\, v) \mid (\empH, v) \in \V{k}{t} \} \\
  \V{k}{ t_1 \otimes t_2 } &= \{ (\varsigma_1 \star \varsigma_2, ( v_1, v_2 )) \mid (\varsigma_1, v_1) \in \V{k}{t_1} \wedge (\varsigma_2, v_2) \in \V{k}{t_2} \}
\end{align*}

The definition of $\V{k}{ '\!f\!c.\ t}$ says a value and heap
must be the same regardless of what fraction is substituted into both; the
$k-1$ is to take into account fraction specialisation takes ones step
(\textsc{Op\_Spc}).

\vspace{-\baselineskip}
\[
    \V{k}{ '\!f\!c.\  t } = \{ (\varsigma, \ottkw{fun}\ '\!f\!c \rightarrow \, v) \mid \forall f.\ (\varsigma [f / f\!c], v [f / f\!c]) \in \V{k-1}{ t [f / f\!c] } \}
\]

@@

In this light, $\V{k}{t' \multimap t}$ simply says
that $v$ is a function and that evaluating the application of it to any
argument (of the correct type, requiring its own set of resources, bounded by
$k$ steps) satisfies all the aforementioned properties.
\begin{align*}
    \V{k}{ t' \multimap t } &= \{ (\varsigma_v, v ) \mid ( v \equiv \ottkw{fun}\ x : t' \rightarrow e \vee v \equiv \ottkw{fix} (g, x : t' , e : t) ) \, \wedge\\
                            & \qquad \qquad \forall j \leq k, (\varsigma_{v'}, v') \in \V{j}{ t' }.\ \varsigma_v \star \varsigma_v' \textrm{ defined } \Rightarrow (\varsigma_v \star \varsigma_v', v\, v') \in \C{j}{t} \}
\end{align*}

The interpretation of typing environments $\Delta$ and $\Gamma$ are with
respect to an arbitrary substitution of fractional permissions $\theta$. Note
that only the interpretation of $\Gamma$ involves a (potentially) non-empty heap.
\begin{align*}
    \den{I}{k}{ \Delta, x : t } \theta &= \{ \delta[x \mapsto v_x] \mid \delta \in \den{I}{k}{\Delta}\theta \wedge (\empH, v_x) \in \V{k}{\theta(t)} \} \\
    \den{L}{k}{ \Gamma, x : t } \theta &= \{ (\varsigma \star \varsigma_x, \gamma[x \mapsto v_x]) \mid (\varsigma, \gamma) \in \den{L}{k}{\Gamma}\theta \wedge (\varsigma_x, v_x) \in \V{k}{\theta(t)} \}
\end{align*}

### 2019-04-13 write-up/paper/intro.tex
Make stuff consistent
Syntax & spacing, with a few small edits from Reviewer C's review.

As a result, programmers often turn to higher-level languages such as
Matlab, R and Numpy, which offer very high-level array abstractions
that can be viewed as ordinary mathematical values. This makes
programming safer, as well as making prototyping and verification much
easier, since it lets programmers write programs which bear a closer
resemblance to the formulas that the mathematicians and statisticians
designing these algorithms prefer to work with, and ensures that
program bugs will reflect incorrectly-computed values rather than heap
corruption.

@@

So in practice, programmers face a tradeoff: they can eschew safety
and exploit the full power of the underlying linear algebra libraries,
or they can obtain safety at the price of unneeded copies and worse
memory efficiency. In this work, show that this tradeoff is not a
fundamental one.

>>>>

As a result, programmers often turn to higher-level languages such as
Matlab, R and NumPy, which offer very high-level array abstractions
that can be viewed as ordinary mathematical values. This makes
programming safer, as well as making prototyping and verification much
easier, since it lets programmers write programs which bear a closer
resemblance to the formulas that the mathematicians and statisticians
designing these algorithms prefer to work with, and ensures that
program bugs will reflect incorrectly-computed values rather than heap
corruption.

@@

So in practice, programmers face a tradeoff: they can eschew safety
and exploit the full power of the underlying linear algebra libraries,
or they can obtain safety at the price of unneeded copies and worse
memory efficiency. In this work, we show that this tradeoff is not a
fundamental one.

### 2019-04-13 write-up/paper/lang_and_examples.tex
Make stuff consistent
Syntax & spacing, with a few small edits from Reviewer C's review.

\subsection{Overview}

The core type theory of \lang\ is a nearly off-the-shelf linear type
theory, supporting familiar features such as linear function spaces
$A \multimap B$ and tensor products $A \otimes B$. We adopt linearity
-- the restriction that each program variable be used at most once --
since it allows us to express purely functional APIs for numerical
library routines that mutate arrays and matrices~\cite{Wadler90}. Due
to linearity, values cannot alias and are only used once, which means
that linearly-typed updates result in no \emph{observable} mutation.

@@

An example of this consuming and re-binding is in \highl{let (row, !x1) =
row[i]}. Indexing is implemented as a primitive \highl{get: 'x. 'x arr --o
!int --o 'x arr * !elt}. Although fractional permissions can be passed around
explicitly  (as done in the recursive call), they can also be
\emph{automatically inferred at call sites}: \highl{row[i] == get _ row i}
takes advantage of this convenience.

@@

Mutating an array is implemented similarly to indexing one: a primitive
\highl{set: z arr --o !int --o !elt --o z arr}.  It consumes the original
array and returns a new array with the updated value.  \highl{let written =
write[i] := <exp> } is just syntactic sugar for \highl{let written = set write i <exp>}.

Since \highl{write: z arr} (where \texttt{z} stands for $k=0$, representing a
fractional permission of $2^{-k} = 2^{-0} = 1$), we may mutate it, but since we
only need to read from \texttt{weights}, its fractional permission index can be
universally-quantified. In the recursive call, we see \texttt{\_} being used
explicitly to tell the compiler to \emph{infer} the correct fractional
permission based on the given arguments.

@@

To square a matrix, first, we extract the dimensions of the argument
\texttt{x}. Then, because we need to use \texttt{x} twice (so that we can
multiply it by itself) but linearity only allows one use, we use \highl{shareM:
'x. 'x mat --o 'x s mat * 'x s mat} to split the permission \texttt{'x} (which
represents $2^{-x}$) into two halves (\texttt{'x s}, which represents $2^{-(x+1)}$).

@@

By using some simple pattern-matching and syntactic sugar, we can:
\begin{itemize}
    \item write normal-looking, apparently non-linear code
    \item use matrix expressions directly and have a call to an efficient
        call to a BLAS/LAPACK routine inserted with appropriate re-bindings
    \item retain the safety of linear types with fractional permissions by
        having the compiler statically enforce the aliasing and read/write rules
        implicitly assumed by BLAS/LAPACK routines.
\end{itemize}

@@

After computing \texttt{x\_T\_x}, we need to invert it and then multiply it by
\texttt{xy}. The BLAS routine \highl{posv: z mat --o z mat --o z mat * z mat}
does exactly that: assuming the first argument is symmetric, \texttt{posv}
mutates its second argument to contain the desired value. Its first argument is
also mutated to contain the (upper triangular) Cholesky decomposition factor of
the original matrix. Since we do not need that matrix (or its memory) again, we
\texttt{free} it. If we forgot to, we would get a \texttt{Variable to\_del not
used} error. Lastly, we return the \texttt{answer} alongside the untouched
input matrices \texttt{(x,y)}.

\subsubsection{L1-Norm Minimisation on Manifolds}

L1-Norm minimisation is often used in optimisation problems, as a
\emph{regularisation} term for reducing the influence of outliers.  Although
the below formuation\cite{bronstein} is intended to be used with \emph{sparse}
computations, \lang's current implementation only implements dense ones.
However, it still serves as a useful example of explaining \lang's features.

@@

Figure~\ref{fig:lang_l1_norm_min} shows even more pattern-matching. Patterns of
the form \highl{let <id> <- [| beta * c + alpha * a * b |]} are also desugared to
\texttt{gemm} calls. Primitives like \highl{transpose: 'x. 'x mat --o 'x mat *
z mat} and \highl{eye: !int --o z mat} allocate new matrices;
\texttt{transpose} returns the transpose of a given matrix and \texttt{eye k}
evaluates to a $k \times k$ identity matrix.

@@

A \emph{Kalman Filter}\cite{kalman} is a an algorithm for combining prior
knowledge of a state, a statistical model and measurements from (noisy) sensors
to produce an estimate a more reliable estimated of the current state.  It has
various applications (navigation, signal-processing, econometrics) and is
relevant here because it is usually presented as a series of complex matrix
equations.

>>>>

\subsection{Syntax}

\lang's concrete syntax is inspired by that of OCaml. It desugars
(Figure~\ref{fig:lang_desugar}) into a smaller, core type and expression
grammar (Figure~\ref{fig:core_grammar}).

\begin{figure}[p]
    \begin{center}
    \[\def\arraystretch{1.3}
    \begin{array}{rcl}
        f & ::= & {'\!f\!c}
                \mid \Zf
                \mid f\ \Sf \\
                \\
        t & ::= & \Unit
                \mid \Bool
                \mid \Int
                \mid \Elt
                \mid f\ \Arr
                \mid f\ \Mat
                \mid \Bang t
                \mid {'\!f\!c}.\ t
                \mid t \otimes t'
                \mid t \multimap t' \\
                \\
        e & ::= & p \textrm{ (primitives)}
                \mid x \textrm{ (variable)}
                \mid \Let\ x = e\ \In\ e'
                \mid ()
                \mid \Let\ () = e\ \In\ e'
                \mid \ottkw{true}
                \mid \ottkw{false}
                \\ && \ottkw{if}\ e\ \ottkw{then}\ e_1\ \ottkw{else}\ e_2
                \mid k \textrm{ (integer)}
                \mid l \textrm{ (heap location)}
                \mid el \textrm{ (array element)}
                \\ && \Many\ v
                \mid \Let\ \Many\ x = e\ \In\ e'
                \mid \ottkw{fun}\ '\!f\!c \rightarrow e
                    \textrm{ (frac. perm. abstraction)}
                \\ && e\ [f] \textrm{ (frac. perm. specialisiation)}
                \mid (e, e')
                \mid \Let\ (a,b) = e\ \In\ e'
                \\ && \ottkw{fun}\ x : t \rightarrow e
                \mid e\ e'
                \mid \ottkw{fix} \: (g, x : t, e : t')
    \end{array} \]
    \end{center}
    \caption{Core fraction $f$, type $t$ and expression $e$ grammar of
        \lang. Values $v$ are a subset of the expressions, their full
        definition and a list of all primitives $p$ is in
        Appendix~\ref{sec:grammar_def}.}\label{fig:core_grammar}
\end{figure}

\begin{figure}[p]
\begin{center}
\[\def\arraystretch{1.3}
    \begin{array}{rcl}
    x[e] &
    \Rightarrow &
    \ottkw{get}\ \_\ x\ (e) \:\qquad \textrm{(similarly for matrices)}
\\
    x[e_1] := e_2 &
    \Rightarrow &
    \ottkw{set}\ x\ (e_1)\ (e_2) \quad \textrm{(similarly for matrices)}
\\
\\
    pat & ::= & () \mid x \mid {!x} \mid \Many\ pat \mid (pat, pat)
\\
    \Let\ !x = e_1\ \In\ e_2 &
    \Rightarrow &
    \specialcell[t]{l}{%
        \Let\ \Many\ x = e_1\ \In \\
        \Let\ \Many\ x = \Many\ (\Many\ x)\ \In\ e_2}
\\
    \Let\ \Many\ \langle pat_x \rangle\ = e_1\ \In\ e_2 &
    \Rightarrow &
    \specialcell[t]{l}{%
        \Let\ \Many\ x = x\ \In \\
        \Let\ \langle pat_x \rangle\ = x\ \In\ e_2}
\\
    \Let\ (\langle pat_a \rangle, \langle pat_b \rangle)\ = e_1\ \In\ e_2 &
    \Rightarrow &
    \specialcell[t]{l}{%
        \Let\ (a,b)\ = a\_b\ \In\
        \Let\ \langle pat_a \rangle\ = a\ \In \\
        \Let\ \langle pat_b \rangle\ = b\ \In\ e_2}
\\
    \ottkw{fun}\ (\langle pat_x \rangle : t) \rightarrow e &
    \Rightarrow &
    \ottkw{fun}\ (x : t) \rightarrow \Let\ \langle pat_x \rangle = x\ \In\ e
\\
\\
    arg & ::= & \langle pat \rangle : t\ \mid\ {'x} \textrm{ (fractional permission variable)}
\\
    \ottkw{fun}\ \langle arg_{1 .. n} \rangle \rightarrow e &
    \Rightarrow &
    \ottkw{fun}\ \langle arg_1 \rangle \rightarrow ..
    \ \ottkw{fun}\ \langle arg_n \rangle \rightarrow e
\\
    \Let\ f\ {\langle arg_{1 .. n} \rangle} = e_1\ \In\ e_2 &
    \Rightarrow &
    \Let\ f = \ottkw{fun}\ {\langle arg_{1 .. n} \rangle} \rightarrow e_1\
    \In\ e_2
\\
    \Let\ !f\ {\langle arg_{1 .. n} \rangle} = e_1\ \In\ e_2 &
    \Rightarrow &
    \Let\ \Many\ f = \Many\ (\ottkw{fun}\ {\langle arg_{1 .. n} \rangle}
    \rightarrow e_1)\ \In\ e_2
\\
    \mathrm{fixpoint} & \equiv & \ottkw{fix}\ (f, x : t, \ottkw{fun}
    \ {\langle arg_{0 .. n} \rangle} \rightarrow e_1 : {t'} )
\\
    \Let\ \ottkw{rec}\ f\ (x : t)\ {\langle arg_{0 .. n} \rangle} : {t'} = e_1\ \In\ e_2 &
    \Rightarrow &
    \Let\ f = \mathrm{fixpoint}\ \In\ e_2
\\
    \Let\ \ottkw{rec}\ !f\ (x : t)\ {\langle arg_{0 .. n} \rangle} : {t'} = e_1\ \In\ e_2 &
    \Rightarrow &
    \Let\ \Many\ f = \Many\ \mathrm{fixpoint}\ \In\ e_2
    \end{array}
\]
\end{center}
\caption{Desugaring \lang.}\label{fig:lang_desugar}
\end{figure}

\subsection{Type System and Other Features}

The core type theory of \lang\ is a nearly off-the-shelf linear type
theory~\cite{barber1996}, supporting familiar features such as linear function
spaces $A \multimap B$ and tensor products $A \otimes B$. We adopt linearity --
the restriction that each program variable be used at most once -- since it
allows us to express purely functional APIs for numerical library routines that
mutate arrays and matrices~\cite{Wadler90}. Due to linearity, values cannot
alias and are only used once, which means that linearly-typed updates result in
no \emph{observable} mutation.

@@

An example of this consuming and re-binding is in \highl{let (row, !x1) =
row[i]}. Indexing is implemented as a primitive $\ottkw{get}:\ {'x}.\ {{'x}\
\Arr} \multimap {!\Int} \multimap {'x\ \Arr} \otimes {!\Elt}$.
Although fractional permissions can be passed around explicitly  (as done in
the recursive call), they can also be \emph{automatically inferred at call
sites}: \highl{row[i] == get _ row i} takes advantage of this convenience.

@@

Mutating an array is implemented similarly to indexing one -- a primitive
$\ottkw{set}:\ {\Zf\ \Arr} \multimap {!\Int} \multimap {!\Elt} \multimap
{\Zf\ \Arr}$. It consumes the original array and returns a new array with
the updated value.  \highl{let written = write[i] := <exp> } is just syntactic
sugar for \highl{let written = set write i <exp>}.

Since \highl{write: z arr} (where \Zf stands for $k=0$, representing a
fractional permission of $2^{-k} = 2^{-0} = 1$), we may mutate it, but since we
only need to read from \texttt{weights}, its fractional permission index can be
universally-quantified. In the recursive call, we see \texttt{\_} being used
explicitly to tell the compiler to \emph{infer} the correct fractional
permission based on the given arguments.

\subsubsection{Digression: Types of Primitives}

\emph{The most pertinent aspect of \lang\ is the types of its primitives}
(Figure~\ref{fig:primitive_types}).  While the types of operations such as
\textbf{get} and \textbf{set} might be borderline obvious, the types of
BLAS/LAPACK routines become an \emph{incredibly useful, automated check for
using the API correctly.}

\begin{figure}
    \begin{center}
    \[\def\arraystretch{1.3}
    \begin{array}{rcl}
%
        \ottkw{symm} &:& {!\Bool} \multimap {!\Elt} \multimap {'x}.\
            {{'x}\ \Mat} \multimap {'y}.\ {{'y}\ \Mat} \multimap {!\Elt}
            \multimap {\Zf\ \Mat} \multimap
            \\ && ( {{'x}\ \Mat} \otimes {{'y}\ \Mat} ) \otimes {\Zf\ \Mat} \\
%
        \ottkw{gemm} &:& {!\Elt} \multimap {'x}.\ {{'x}\ \Mat} \otimes
            {!\Bool} \multimap {'y}.\ {{'y}\ \Mat} \otimes {!\Bool}
            \multimap {!\Elt} \multimap
            \\ && {\Zf\ \Mat} \multimap ( {{'x}\ \Mat} \otimes {{'y}\ \Mat} )
            \otimes {\Zf\ \Mat} \\
%
        \ottkw{gesv} &:& {\Zf\ \Mat} \multimap {\Zf\ \Mat} \multimap
            {\Zf\ \Mat} \otimes {\Zf\ \Mat} \\
%
        \ottkw{posv} &:& {\Zf\ \Mat} \multimap {\Zf\ \Mat} \multimap
            {\Zf\ \Mat} \otimes {\Zf\ \Mat} \\
%
        \ottkw{potrs} &:& {'x}.\ {{'x}\ \Mat} \multimap {\Zf\ \Mat}
            \multimap {{'x}\ \Mat} \otimes {\Zf\ \Mat} \\
%
        \ottkw{syrk} &:& {!\Bool} \multimap {!\Elt} \multimap {'x}.\
            {{'x}\ \Mat} \multimap {!\Elt} \multimap {\Zf\ \Mat} \multimap
            {{'x}\ \Mat} \otimes {\Zf\ \Mat} \\
%
    \end{array} \]
    \end{center}
    \caption{Types of some \lang\ primitives.}\label{fig:primitive_types}
\end{figure}

We determine the types for these routines by consulting their documentation.
Each routine has a record of the expected aliasing behaviour and whether or not
it modifies or consumes its argument in any way. We use that to derive the
types in Figure~\ref{fig:primitive_types}. Since most of these low-level
routines are very careful not to do any allocation themselves, it is generally
very easy to give each a \lang\ type -- every argument that can modify/consume
its argument needs a full permission, and all others can be
fraction-polymorphic.  Taking Fortran as an example, it has a notion of
\texttt{in}, \texttt{out} and \texttt{inout} parameters. The latter two would
need full \Zf permissions; the first would be fraction-polymorphic.

@@

To square a matrix, first, we extract the dimensions of the argument
\texttt{x}. Then, because we need to use \texttt{x} twice (so that we can
multiply it by itself) but linearity only allows one use, we use
$\ottkw{shareM}:\ {'x}.\ {{'x}\ \Mat} \multimap {{'x}\ \Sf\ \Mat} \otimes
{{'x}\ \Sf\ \Mat}$ to split the permission \texttt{'x} (which represents
$2^{-x}$) into two halves (\texttt{'x s}, which represents $2^{-(x+1)}$).

@@

By using some simple pattern-matching and syntactic sugar
(Figure~\ref{fig:lang_matexp}), we can:
\begin{itemize}
    \item write normal-looking, apparently non-linear code
    \item use matrix expressions directly and have a call to an efficient
        call to a BLAS/LAPACK routine inserted with appropriate re-bindings
    \item retain the safety of linear types with fractional permissions by
        having the compiler statically enforce the aliasing and read/write rules
        implicitly assumed by BLAS/LAPACK routines.
\end{itemize}

\begin{figure}[t]
\begin{center}
\[\def\arraystretch{1.3}
    \begin{array}{rcl}
    \Let\ v \leftarrow x[e]\ \In\ e &
    \Rightarrow &
    \Let\ (x, !v)\ = x[e]\ \In\ e \qquad \textrm{(similarly for matrices)}
\\
    \Let\ x_2 \leftarrow \ottkw{new}\ [|\ x_1\ |]\ \In\ e &
    \Rightarrow &
    \Let\ (x_1, x_2)\ = \ottkw{copyM}\ \_\ x_1\ \In\ e
\\
    \Let\ x_2 \leftarrow [|\ x_1\ |]\ \In\ e &
    \Rightarrow &
    \Let\ (x_1, x_2)\ = \ottkw{copyM\_to}\ \_\ x_1\ x_2\ \In\ e
    \end{array}
\]
\[
    M ::= X\ \mid\ X^T\ \mid\ \textrm{sym}( X )
\]
\end{center}
\begin{align*}
% new
    \Let\ &Y \leftarrow \ottkw{new}\ (n, k)\ [|\ \alpha M_1 M_2 \ |]\ \In\ e
    \Rightarrow \\
    & \Let\ Y\ = \ottkw{matrix}\ n\ k\ \In
    \ \Let\ Y \leftarrow [|\ \alpha M_1 M_2 + 0Y\ |]\ \In\ e
\\[0.1\baselineskip]
%syrk
    \Let\ &Y \leftarrow [|\ \alpha X X^T + \beta Y\ |]\ \In\ e
    \Rightarrow \\
    & \Let\ (X,Y)\ = \ottkw{syrk}\ \ottkw{false}
    \ \alpha\ \_\ X\ \beta\ Y\ \In\ e
\\[0.1\baselineskip]
    \Let\ &Y \leftarrow [|\ \alpha X^T X + \beta Y\ |]\ \In\ e
    \Rightarrow \\
    & \Let\ (X,Y)\ = \ottkw{syrk}\ \ottkw{true}
    \ \alpha\ \_\ X\ \beta\ Y\ \In\ e
%symm
\\[0.1\baselineskip]
    \Let\ &Y \leftarrow\ [|\ \alpha\,\textrm{sym}(X_1)\,X_2 + \beta Y\ |]\ \In\ e
    \Rightarrow \\
    & \Let\ ((X_1,X_2),Y)\ = \ottkw{symm}\ \ottkw{false}
        \ \alpha\ \_ \ X_1 \_\ X_2\ \beta\ Y\ \In\ e
\\[0.1\baselineskip]
    \Let\ &Y \leftarrow\ [|\ \alpha\,X_2\,\textrm{sym}(X_1) + \beta Y\ |]\ \In\ e
    \Rightarrow \\
    & \Let\ ((X_1,X_2),Y)\ = \ottkw{symm}\ \ottkw{true}
        \ \alpha\ \_ \ X_1 \_\ X_2\ \beta\ Y\ \In\ e
\\[0.1\baselineskip]
%gemm
    \Let\ &Y \leftarrow [|\ \alpha X_1^{T?} X_2^{T?} + \beta Y\ |]\ \In\ e
    \Rightarrow \\
    & \Let\ ((X_1,X_2),Y)\ =\ \ottkw{gemm}\ \alpha
        \ \_\ \left(X_1, \substack{\ottkw{true}\\\ottkw{false}}\right)
        \ \_\ \left(X_2, \substack{\ottkw{true}\\\ottkw{false}}\right)
        \ \beta\ Y\ \In\ e
\end{align*}

\caption{Purely syntactic pattern-matching translations of
    matrix expressions.}\label{fig:lang_matexp}
\end{figure}

@@

After computing \texttt{x\_T\_x}, we need to invert it and then multiply it by
\texttt{xy}. The BLAS routine $\ottkw{posv}:\ {\Zf\ \Mat} \multimap
{\Zf\ \Mat} \multimap {\Zf\ \Mat} \otimes {\Zf\ \Mat}$ does
exactly that: assuming the first argument is symmetric, \texttt{posv} mutates
its second argument to contain the desired value. Its first argument is also
mutated to contain the (upper triangular) Cholesky decomposition factor of the
original matrix. Since we do not need that matrix (or its memory) again, we
\texttt{free} it. If we forgot to, we would get a \texttt{Variable to\_del not
used} error. Lastly, we return the \texttt{answer} alongside the untouched
input matrices \texttt{(x,y)}.

\subsubsection{L1-norm Minimisation on Manifolds}

L1-norm minimisation is often used in optimisation problems, as a
\emph{regularisation} term for reducing the influence of outliers.  Although
the below formulation~\cite{bronstein} is intended to be used with \emph{sparse}
computations, \lang's current implementation only implements dense ones.
However, it still serves as a useful example of explaining \lang's features.

@@

Figure~\ref{fig:lang_l1_norm_min} shows even more pattern-matching. Patterns of
the form \highl{let <id> <- [| beta * c + alpha * a * b |]} are also desugared
to \texttt{gemm} calls. Primitives like $\ottkw{transpose}:\ {'x}.\ {{'x}\ \Mat}
\multimap {{'x}\ \Mat} \otimes {\Zf\ \Mat}$ and $\ottkw{eye}:\ {!\Int}
\multimap {\Zf\ \Mat}$ allocate new matrices; \texttt{transpose} returns
the transpose of a given matrix and \texttt{eye k} evaluates to a $k \times k$
identity matrix.

@@

A \emph{Kalman Filter}~\cite{kalman} is an algorithm for combining prior
knowledge of a state, a statistical model and measurements from (noisy) sensors
to produce an estimate a more reliable estimated of the current state.  It has
various applications (navigation, signal-processing, econometrics) and is
relevant here because it is usually presented as a series of complex matrix
equations.
